% =============================================================================
% \section{Methodology}
% NeurIPS style -- compile with pdflatex + amsmath + amssymb + algorithm2e
% =============================================================================

\section{Methodology}
\label{sec:methodology}

We propose \textbf{ForeBlocks-DARTS}, a differentiable neural architecture
search framework for multivariate time series forecasting. Given a look-back
window of length $L$ and input dimensionality $C$, the model predicts a
forecast horizon of length $H$. The search space spans (i) the topology and
operator mixture inside a DARTS cell, (ii) the backbone topology mode
(encoder--decoder, encoder-only, decoder-only, or Mamba), (iii) the internal
configuration of transformer encoder/decoder blocks, including tokenizer,
attention kernel, positional bias, feed-forward variant, and decoding style,
and (iv) the input normalization strategy. All search variables are optimized
within a differentiable bi-level formulation.

% ---------------------------------------------------------------------------
\subsection{Problem Formulation}
\label{sec:problem}

Let $\mathbf{X} \in \mathbb{R}^{B \times L \times C}$ denote a batch of
multivariate time series and let
$\hat{\mathbf{Y}} \in \mathbb{R}^{B \times H \times C}$ denote the forecast.
We seek a parametric map
$f_{\boldsymbol{\alpha},\boldsymbol{w}} : \mathbb{R}^{L \times C}
\rightarrow \mathbb{R}^{H \times C}$ that minimizes prediction error on
held-out data. Following DARTS~\citep{liu2019darts}, the learnable variables
are partitioned into architecture parameters $\boldsymbol{\alpha}$ and model
weights $\boldsymbol{w}$. The bi-level objective is
%
\begin{align}
    \min_{\boldsymbol{\alpha}} \quad
    & \mathcal{L}_{\mathrm{val}}\!\left(
        \boldsymbol{w}^*(\boldsymbol{\alpha}),\, \boldsymbol{\alpha}
      \right)
      + \lambda_{\mathrm{eff}}\,\Omega_{\mathrm{eff}}(\boldsymbol{\alpha})
      + \lambda_{\beta}\,\lVert \boldsymbol{\alpha} \rVert_2^2
      + \lambda_{\mathrm{moe}}\,\Omega_{\mathrm{moe}}(\boldsymbol{\alpha})
      - \lambda_{\mathrm{exp}}\,\Omega_{\mathrm{exp}}(\boldsymbol{\alpha})
    \label{eq:outer} \\
    \text{s.t.} \quad
    & \boldsymbol{w}^*(\boldsymbol{\alpha}) =
      \arg\min_{\boldsymbol{w}}\;
      \mathcal{L}_{\mathrm{train}}(\boldsymbol{w}, \boldsymbol{\alpha}),
    \label{eq:inner}
\end{align}
%
where $\mathcal{L}$ is a Huber loss, $\Omega_{\mathrm{eff}}$ is a
FLOPs-normalized efficiency penalty, the $\ell_2$ term corresponds to the
$\beta$-DARTS regularizer~\citep{ye2022betadarts},
$\Omega_{\mathrm{moe}}$ is a lightweight expert-balance penalty for the MoE
feed-forward branch, and $\Omega_{\mathrm{exp}}$ is an early-stage entropy
bonus applied to transformer selection logits to reduce premature collapse to
a single attention or feed-forward mode.

% ---------------------------------------------------------------------------
\subsection{Input Normalization Search}
\label{sec:norm}

Time series frequently exhibit non-stationarity and distribution shift. We
therefore use a searchable normalization head that interpolates among
Reversible Instance Normalization (RevIN)~\citep{kim2021reversible},
InstanceNorm, and an identity mapping. Let
$\boldsymbol{\alpha}^{\mathrm{norm}} \in \mathbb{R}^{3}$ denote the
normalization logits. The transformed input is
%
\begin{equation}
    \tilde{\mathbf{X}} =
      w_1\,\mathrm{RevIN}(\mathbf{X}) +
      w_2\,\mathrm{InstanceNorm}(\mathbf{X}) +
      w_3\,\mathbf{X},
    \qquad
    \mathbf{w} = \mathrm{softmax}\!\left(
      \tfrac{\boldsymbol{\alpha}^{\mathrm{norm}}}{\tau}\right).
    \label{eq:norm}
\end{equation}
%
At inference, the argmax-selected transform is applied as a hard choice. When
RevIN is selected, the inverse transformation is applied to the forecast to
restore the original scale.

% ---------------------------------------------------------------------------
\subsection{DARTS Cells and Cell-Level Operator Search}
\label{sec:cell}

\paragraph{Cell DAG.}
Each cell is a directed acyclic graph with ordered nodes
$\{\mathbf{h}_0, \ldots, \mathbf{h}_{N-1}\}$. Node $i$ aggregates all
preceding nodes $j < i$ through edge-specific mixed operations
$\bar{o}_{j,i}$. The update rule is
%
\begin{equation}
    \mathbf{h}_i
    = \sigma(r_i)
      \Bigg(\sum_{j < i} \sigma(\gamma_{ji})\,
        \bar{o}_{j,i}(\mathbf{h}_j)
      \Bigg)
      + \bigl(1 - \sigma(r_i)\bigr)\,\mathbf{h}_{i-1},
    \label{eq:cell}
\end{equation}
%
where $r_i$ is a residual gate and $\gamma_{ji}$ is an edge-importance logit.
The output of the cell is the normalized representation of the final node.

\paragraph{Mixed operation.}
For a selected candidate operator pool $\mathcal{O}$, each edge computes a
weighted mixture
%
\begin{equation}
    \bar{o}(\mathbf{x}) = \sum_{k \in \mathcal{O}} p_k\,o_k(\mathbf{x}),
    \qquad
    p_k = \frac{\phi(z_k)}{\sum_{\ell}\phi(z_{\ell})},
    \label{eq:mixedop}
\end{equation}
%
where $z_k$ are learnable edge logits and $\phi(\cdot)$ denotes the
competition-aware relaxation used during search. The implementation supports
single-path routing, Fair DARTS gating~\citep{chu2021fair}, and the
Dirichlet relaxation of DrNAS~\citep{chen2021drnas}.

\paragraph{Hierarchical operator pool construction.}
The differentiable edge mixture is applied over a candidate pool that is first
constructed hierarchically in the outer multi-fidelity search. Candidate cell
operators are sampled by family and then by operation within family. The
families are convolutional, frequency-domain, attention-inspired, and
MLP-style transformations, while the state-space family is represented by the
separate Mamba topology rather than as a cell operator. Identity is retained
as a residual-preserving operator. Table~\ref{tab:ops} lists the 16 available
cell operators.

\paragraph{Temperature annealing.}
Architecture temperatures follow a cosine schedule,
%
\begin{equation}
    \tau(e) = \tau_{\min} + \tfrac{1}{2}(\tau_0 - \tau_{\min})
      \left(1 + \cos\!\left(\tfrac{e - e_w}{E - e_w}\pi\right)\right),
    \label{eq:temp}
\end{equation}
%
where $e$ is the current epoch, $E$ is the total number of epochs, and $e_w$
is the warm-up period. Annealing drives the relaxed distributions from broad
exploration toward near-discrete selections.

\begin{table}[h]
\centering
\caption{Cell-level candidate operations ($|\mathcal{O}|=16$). All operators
         map $\mathbb{R}^{L \times d_l}$ to $\mathbb{R}^{L \times d_l}$.}
\label{tab:ops}
\small
\begin{tabular}{lll}
\toprule
Family & Operation & Inductive bias \\
\midrule
Residual
  & Identity & residual shortcut \\
\addlinespace
Convolutional
  & TimeConv & local temporal patterns \\
  & TCN & dilated temporal convolution \\
  & ConvMixer & depthwise-pointwise token mixing \\
  & MultiScaleConv & multi-kernel aggregation \\
  & PyramidConv & hierarchical temporal pyramid \\
\addlinespace
Frequency
  & Wavelet & multi-resolution decomposition \\
  & Fourier & global spectral mixing \\
  & DLinear & trend-seasonality linear decomposition \\
  & TimesNet & temporal-to-2D periodic modeling \\
\addlinespace
Attention-inspired
  & PatchEmbed & patch tokenization \\
  & InvertedAttention & channel-wise token mixing \\
\addlinespace
MLP-style
  & GRN & gated residual transformation \\
  & ResidualMLP & position-wise feed-forward mapping \\
  & TimeMixer & multiscale mixing block \\
  & NBeats & basis-expansion forecasting block \\
\bottomrule
\end{tabular}
\end{table}

% ---------------------------------------------------------------------------
\subsection{Backbone and Feature Fusion}
\label{sec:backbone}

The normalized input is projected to hidden dimension $d_h$ by a linear
embedding layer followed by LayerNorm and GELU. A stack of $K$ DARTS cells
then produces intermediate feature maps
%
\begin{equation}
    \mathbf{F}^{(k)} = \mathrm{Cell}^{(k)}\!\left(\mathbf{F}^{(k-1)}\right),
    \qquad
    \mathbf{F}^{(0)} = \mathrm{Embed}(\tilde{\mathbf{X}}).
    \label{eq:cells}
\end{equation}
%
Cell outputs are combined with learned cell importances and cell-combination
weights,
%
\begin{equation}
    \mathbf{F} = \sum_{k=1}^{K}
      \frac{\sigma(\gamma_k)\,\bar{w}_k}
           {\sum_j \sigma(\gamma_j)\,\bar{w}_j}
      \mathbf{F}^{(k)},
    \label{eq:cell_comb}
\end{equation}
%
followed by gated residual fusion with the input embedding,
%
\begin{equation}
    \mathbf{F}^* =
      \mathbf{g} \odot \mathbf{F} +
      (1 - \mathbf{g}) \odot \mathbf{F}^{(0)} +
      \beta_s\,\mathbf{F}^{(0)},
    \label{eq:gate}
\end{equation}
%
where $\mathbf{g}$ is a learnable gate and $\beta_s$ is a global skip weight.
The fused representation $\mathbf{F}^*$ is passed to the selected backbone
mode.

% ---------------------------------------------------------------------------
\subsection{Topology Modes}
\label{sec:topologies}

The architecture operates in one of four topology modes,
$m \in \{\texttt{encoder\_decoder},\texttt{encoder\_only},
\texttt{decoder\_only},\texttt{mamba}\}$.

\paragraph{(A) Encoder--Decoder.}
The default sequence model is a transformer encoder--decoder. The encoder
maps $\mathbf{F}^*$ to an encoded sequence
$\mathbf{H}_{\mathrm{enc}} \in \mathbb{R}^{B \times L \times d_l}$, a
context token $\mathbf{c}$, and a latent state summary. The decoder then
produces the horizon either autoregressively or in a parallel Informer-style
manner, both conditioned on $\mathbf{H}_{\mathrm{enc}}$.

\paragraph{(B) Encoder-Only.}
The decoder is removed. The transformer encoder output is pooled over time and
blended with the last token, after which a direct projection head maps the
representation to $\mathbb{R}^{H \times C}$.

\paragraph{(C) Decoder-Only.}
The encoder is removed. The cell backbone is projected into decoder memory,
and the transformer decoder generates the forecast conditioned only on this
compressed backbone context.

\paragraph{(D) Mamba SSM.}
Both encoder and decoder are replaced by a stack of selective state-space
(Mamba) layers~\citep{gu2023mamba}. This branch operates directly on the
backbone representation and produces a non-autoregressive forecast through a
projection head.

% ---------------------------------------------------------------------------
\subsection{Searchable Transformer Components}
\label{sec:transformers}

The encoder and decoder are transformer-first modules. Rather than searching
over recurrent cell types, ForeBlocks-DARTS searches the internal
configuration of the transformer blocks.

\paragraph{Encoder tokenizer.}
The encoder owns a differentiable tokenizer choice among
$\{\texttt{direct},\texttt{patch\_8},\texttt{patch\_16},
\texttt{patch\_32},\texttt{multi\_scale\_patch},
\texttt{variate\_tokens}\}$. Let
$\boldsymbol{\alpha}^{\mathrm{tok}} \in \mathbb{R}^{6}$ denote the tokenizer
logits. The token sequence is
%
\begin{equation}
    \mathbf{T}_{\mathrm{enc}} = \sum_{s=1}^{6}
      p^{\mathrm{tok}}_s\,\mathcal{T}_s(\mathbf{F}^*),
    \qquad
    \mathbf{p}^{\mathrm{tok}} =
    \mathrm{softmax}\!\left(\tfrac{\boldsymbol{\alpha}^{\mathrm{tok}}}{\tau}\right).
    \label{eq:tokenizer}
\end{equation}
%
This formulation unifies direct tokenization, fixed patch lengths, a
multi-scale patch encoder, and a variate-token pathway inspired by
channel-first forecasting models.

\paragraph{Self-attention kernel search.}
Encoder and decoder self-attention layers each maintain architecture logits
over the candidate kernels
$\{\texttt{sdp},\texttt{linear},\texttt{probsparse},\texttt{cosine},
\texttt{local}\}$. The relaxed self-attention operator is
%
\begin{equation}
    \mathcal{A}_{\mathrm{self}}(\mathbf{Q},\mathbf{K},\mathbf{V})
    = \sum_{m=1}^{5} p^{\mathrm{attn}}_m\,
      \mathcal{A}_m(\mathbf{Q},\mathbf{K},\mathbf{V}),
    \label{eq:selfattn_mix}
\end{equation}
%
with $\mathbf{p}^{\mathrm{attn}} =
\mathrm{softmax}(\boldsymbol{\alpha}^{\mathrm{attn}}/\tau)$. In the final
architecture, the argmax-selected kernel is frozen as a discrete choice.

\paragraph{Positional bias search.}
Each self-attention and decoder cross-attention block additionally searches
over the positional treatment
$\{\texttt{rope},\texttt{alibi},\texttt{none},\texttt{seasonal}\}$. This
allows the model to adapt between rotary phase encoding, additive linear bias,
no explicit positional bias, and a seasonal relative-bias parameterization.

\paragraph{Feed-forward search.}
Each transformer layer chooses between a dense SwiGLU feed-forward network and
a simplified sparse mixture-of-experts (MoE) block. With logits
$\boldsymbol{\alpha}^{\mathrm{ffn}} \in \mathbb{R}^{2}$, the feed-forward
output is
%
\begin{equation}
    \mathrm{FFN}(\mathbf{x}) =
      p^{\mathrm{ffn}}_{\mathrm{dense}}\,\mathrm{SwiGLU}(\mathbf{x}) +
      p^{\mathrm{ffn}}_{\mathrm{moe}}\,\mathrm{MoE}(\mathbf{x}),
    \label{eq:ffn_mix}
\end{equation}
%
where the MoE branch uses a lightweight top-$k$ token router over routed
experts and a balance penalty to discourage expert collapse.

% ---------------------------------------------------------------------------
\subsection{Decoder Cross-Attention and Query Generation}
\label{sec:bridge}

The decoder no longer uses a separate external bridge between encoder and
decoder. Instead, searchable cross-attention resides directly inside the
transformer decoder layers. The decoder cross-attention operator searches over
$\{\texttt{none},\texttt{sdp},\texttt{linear},\texttt{probsparse},
\texttt{cosine},\texttt{local}\}$ by the same relaxation as in
Eq.~\eqref{eq:selfattn_mix}. Cross-attention positional bias is searched by
an independent set of logits.

\paragraph{Decoder style.}
The decoder style is chosen differentiably between an autoregressive pathway
and a non-autoregressive Informer-style pathway,
%
\begin{equation}
    \hat{\mathbf{Y}} =
      p^{\mathrm{style}}_{\mathrm{ar}}\,
      \mathcal{D}_{\mathrm{ar}}(\mathbf{X}, \mathbf{H}_{\mathrm{enc}})
      +
      p^{\mathrm{style}}_{\mathrm{inf}}\,
      \mathcal{D}_{\mathrm{inf}}(\mathbf{X}, \mathbf{H}_{\mathrm{enc}}),
    \label{eq:decoder_style}
\end{equation}
%
where $\mathcal{D}_{\mathrm{ar}}$ generates one step at a time and
$\mathcal{D}_{\mathrm{inf}}$ predicts the full horizon in parallel.

\paragraph{Parallel decoder query generator.}
For the Informer-style decoder, the horizon query sequence is itself
searchable. The model chooses among
$\{\texttt{repeat\_last},\texttt{zeros},
\texttt{learned\_horizon\_queries},\texttt{shifted\_target},
\texttt{future\_covariate\_queries}\}$,
%
\begin{equation}
    \mathbf{Q}_{\mathrm{dec}} = \sum_{g} p^{\mathrm{qry}}_g\,
      \mathcal{Q}_g(\mathbf{X}, \mathbf{X}_{\mathrm{future}}),
    \label{eq:query_mix}
\end{equation}
%
where $\mathbf{X}_{\mathrm{future}}$ denotes optional future covariates or
teacher-forcing targets during training.

\paragraph{Memory compression.}
Before decoder cross-attention, the memory sequence may be compressed by a
learned pooling module with a searchable number of memory queries. This
provides a lightweight mechanism for adapting the decoder memory budget
without changing the encoder itself.

% ---------------------------------------------------------------------------
\subsection{Bi-Level Optimization}
\label{sec:bilevel}

The bi-level problem in Eqs.~\eqref{eq:outer}--\eqref{eq:inner} is solved with
alternating optimization. Model weights are updated on the training split,
whereas architecture parameters are updated on a validation split:
%
\begin{enumerate}[nosep]
    \item \textbf{Weight update:}
          $\boldsymbol{w} \leftarrow \boldsymbol{w} -
          \eta_w\,\nabla_{\boldsymbol{w}}
          \mathcal{L}_{\mathrm{train}}(\boldsymbol{w},\boldsymbol{\alpha})$.
    \item \textbf{Architecture update:}
          $\boldsymbol{\alpha} \leftarrow \boldsymbol{\alpha} -
          \eta_{\alpha}\,\nabla_{\boldsymbol{\alpha}}
          \mathcal{L}_{\mathrm{val}}(\boldsymbol{w},\boldsymbol{\alpha})$.
\end{enumerate}
%
Both updates use AdamW. The implementation additionally supports architecture
gradient EMA, progressive shrinking, hybrid pruning, edge diversity
regularization, late-stage entropy sharpening, and the transformer-specific
exploration bonus introduced in Eq.~\eqref{eq:outer}. The overall objective is
%
\begin{equation}
    \mathcal{L}_{\mathrm{total}} =
      \mathcal{L}_{\mathrm{Huber}}
      + \lambda_{\mathrm{KL}}\,\Omega_{\mathrm{KL}}
      + \lambda_{\mathrm{eff}}\,\Omega_{\mathrm{eff}}
      + \lambda_{\beta}\,\lVert \boldsymbol{\alpha} \rVert_2^2
      + \lambda_{\mathrm{moe}}\,\Omega_{\mathrm{moe}}
      - \lambda_{\mathrm{exp}}\,\Omega_{\mathrm{exp}}.
    \label{eq:total_loss}
\end{equation}

% ---------------------------------------------------------------------------
\subsection{Multi-Fidelity Search Pipeline}
\label{sec:multifidelity}

In addition to inner-loop differentiable optimization, the system uses an
outer multi-fidelity search pipeline to compare candidate search spaces and
model scales. First, a candidate pool is generated over hidden dimension,
cell count, node count, topology mode, and operator-family subset. Phase 1
uses zero-cost and proxy metrics, followed by pool-relative rescoring rather
than a fixed global hand-tuned proxy aggregation. Surviving candidates then
undergo short DARTS training under an ASHA-style successive halving schedule:
small budgets are allocated to all candidates, and only the top-performing
subset at each rung is promoted to larger budgets. The final candidate is
converted to a discrete architecture and retrained from scratch.

% ---------------------------------------------------------------------------
\subsection{Architecture Discretization}
\label{sec:discretisation}

After convergence, the relaxed architecture is discretized by selecting the
argmax operator on each cell edge above threshold $\vartheta$,
%
\begin{equation}
    o^*_{j,i} = \arg\max_k\; p_{j,i,k},
    \qquad
    \text{retain iff } p_{j,i,k} > \vartheta.
    \label{eq:discretise}
\end{equation}
%
The same argmax procedure is applied to normalization, encoder tokenizer,
self-attention kernel, positional bias, feed-forward mode, decoder style,
query generator, memory-query count, and decoder cross-attention choices. The
resulting fixed architecture is then retrained on the full training set.

% ---------------------------------------------------------------------------
\subsection{Automatic Mixed Precision and Memory Efficiency}
\label{sec:amp}

All forward and backward passes are executed under PyTorch AMP, reducing
memory consumption relative to full \texttt{float32} training. DARTS cells and
transformer blocks also support gradient checkpointing, which trades
recomputation for memory on deeper search configurations. Architecture
parameters remain in \texttt{float32} to preserve gradient fidelity.

% ---------------------------------------------------------------------------
% References (add to your .bib file)
% ---------------------------------------------------------------------------
%
% @inproceedings{liu2019darts,
%   title={DARTS: Differentiable Architecture Search},
%   author={Liu, H. and Simonyan, K. and Yang, Y.},
%   booktitle={ICLR}, year={2019}}
%
% @inproceedings{ye2022betadarts,
%   title={$\beta$-DARTS: Beta-Decay Regularization for Differentiable Architecture Search},
%   author={Ye, P. and Li, B. and Li, Y. and Chen, T. and Fan, J.},
%   booktitle={CVPR}, year={2022}}
%
% @inproceedings{chen2021drnas,
%   title={DrNAS: Dirichlet Neural Architecture Search},
%   author={Chen, X. and Wang, R. and Cheng, M. and Tang, X. and Heng, P.-A.},
%   booktitle={ICLR}, year={2021}}
%
% @inproceedings{chu2021fair,
%   title={Fair DARTS: Eliminating Unfair Advantages in Differentiable Architecture Search},
%   author={Chu, X. and Zhou, T. and Zhang, B. and Li, J.},
%   booktitle={ECCV}, year={2021}}
%
% @inproceedings{kim2021reversible,
%   title={Reversible Instance Normalization for Accurate Time-Series Forecasting Against Distribution Shift},
%   author={Kim, T. and Kim, J. and Tae, Y. and Park, C. and Choi, J.-H. and Choo, J.},
%   booktitle={ICLR}, year={2022}}
%
% @inproceedings{gu2023mamba,
%   title={Mamba: Linear-Time Sequence Modeling with Selective State Spaces},
%   author={Gu, A. and Dao, T.},
%   booktitle={ICLR}, year={2024}}
%
% @inproceedings{choromanski2020rethinking,
%   title={Rethinking Attention with Performers},
%   author={Choromanski, K. and others},
%   booktitle={ICLR}, year={2021}}
%
% @inproceedings{zhou2021informer,
%   title={Informer: Beyond Efficient Transformer for Long Sequence Time-Series Forecasting},
%   author={Zhou, H. and others},
%   booktitle={AAAI}, year={2021}}
%
% @inproceedings{qin2022cosformer,
%   title={CosFormer: Rethinking Softmax in Attention},
%   author={Qin, Z. and Sun, W. and Deng, H. and Li, D. and Wei, Y. and Lv, B. and Yan, J. and Kong, L. and Zhong, Y.},
%   booktitle={ICLR}, year={2022}}
%
% @inproceedings{cai2020once,
%   title={Once for All: Train One Network and Specialize it for Efficient Deployment},
%   author={Cai, H. and Gan, C. and Wang, T. and Zhang, Z. and Han, S.},
%   booktitle={ICLR}, year={2020}}
%
% @article{chen2016training,
%   title={Training Deep Nets with Sublinear Memory Cost},
%   author={Chen, T. and Xu, B. and Zhang, C. and Guestrin, C.},
%   journal={arXiv:1604.06174}, year={2016}}
